### 2023-01-05 draft.tex
Deploying to gh-pages from @ gottacatchenall/ms_false_negatives@4877591b889630225f889edd6888ba8815a663e0 🚀

Species interactions drive many processes in evolutionary biology and
community ecology. A better understanding of interactions among species
is an imperative to both mitigate the potentially harmful impacts of
anthropogenic change on Earth's biodiversity (Makiola \emph{et al.}
2020) and to predict zoonotic spillover of disease to prevent future
pandemics (Becker \emph{et al.} 2021). However, meeting these challenges
is difficult because interactions are intrinsically hard to sample
(Jordano 2016). Over the past few decades biodiversity data has become
increasingly available---for example, remote-sensing has enabled
collection of data on spatial scales and resolutions previously
unimaginable (Stephenson 2020), while the adoption of open data
practices (Kenall \emph{et al.} 2014) have substantially increased the
amount of data available to ecologists. Still, widespread data about
species interactions remains elusive (Poisot \emph{et al.} 2021).
Observing an interaction between two species often requires human
observation because remote sampling methods can primarily detect
co-occurrence (but see Niedballa \emph{et al.} (2019)), and this itself
is not necessarily indicative of an interaction (Blanchet \emph{et al.}
2020). This constraint induces biases on species interaction data
subject to the spatial and temporal scales that current observation
methods can feasibly sample. This is further compounded by semantic
confusion around the word ``interaction''---for example one might
consider competition a type of species interaction, even though it is
marked by a lack of co-occurrence between species, unlike other types of
interactions, like trophism or pollination, which require both species
to be together at the same place and time. We define interaction in the
latter sense, where two species have fitness consequences on one-another
if they are in the sample place at the same time. In addition, here we
only consider direct (not higher-order) interactions.

The importance of \emph{sampling effort} and its impact on resulting
ecological data has produced a rich body of literature. The recorded
number of species in a dataset or sample depends on the total number of
observations (Walther \emph{et al.} 1995; Willott 2001)---as do
estimates of population abundance (Griffiths 1998)---in addition to
spatial coverage and species detectability. This has motivated more
quantitatively robust approaches to account for error in sampling data
in many contexts: to determine if a given species is extinct (Boakes
\emph{et al.} 2015), to determine sampling design (Moore \& McCarthy
2016), and to measure species richness across large scales (Carlson
\emph{et al.} 2020). In the context of interactions, an initial concern
was the compounding effects of limited sampling effort combined with the
amalgamation of data (across both study sites, time of year, and
taxonomic scales) could lead any empirical set of observations to
inadequately reflect the reality of how species interact (Paine 1988) or
the structure of the network as a whole (Martinez \emph{et al.} 1999;
McLeod \emph{et al.} 2021). Martinez \emph{et al.} (1999) showed that in
a plant-endophyte trophic network, network connectance is robust to
sampling effort, but this was done in the context of a system for which
observation of 62,000 total interactions derived from 164,000
plant-stems was feasible. In some systems (e.g.~megafauna food-webs)
this many observations is either impractical or infeasible due to the
absolute abundance of the species in question.

@@

@@

Here, we illustrate how our confidence that a pair of species never
interacts highly depends on sampling effort, and suggest that surveys of
species interactions can benefit from simulation modeling of the
sampling process. We demonstrate that the realized false-negative rate
of interactions is directly related to the relative abundance of the
species pool, and demonstrate how simulation can be used to produce a
null estimate of the false-negative rate given total sampling effort
(the total count of all individuals of all species seen). We show that
positive associations in co-occurrence data can increase the realized
number of false-negatives, and demonstrate these positive associations
are rampant in network datasets. We conclude by recommending that the
simulation of sampling effort and species occurrence can and should be
used to help design surveys of species interaction diversity (Moore \&
McCarthy 2016), and by advocating use of null models like those
presented here as a tool for both guiding design of surveys of species
interactions and for modeling detection error in predictive models.

@@

\hypertarget{resampling-interaction-probabilities-to-account-for-detection-error}{%
\section{Resampling interaction probabilities to account for detection
error}\label{resampling-interaction-probabilities-to-account-for-detection-error}}

\begin{figure}
\centering
\includegraphics{./figures/uncertainty_sampler.png}
\caption{todo}
\end{figure}

@@

This model above doesn't consider the possibility that there are
positive or negative associations which shift the probability of
observing two species together due to their interaction (Cazelles
\emph{et al.} 2016). However, here we demonstrate that the probability
of observing a false negative can be higher if there is some positive
association in the occurrence of species \(A\) and \(B\). If we denote
the probability that we observe the co-occurrence of two species \(A\)
and \(B\) as \(P(AB)\) and if there is no association between the
marginal probabilities of observing \(A\) and observing \(B\), denoted
\(P(A)\) and \(P(B)\) respectively, then the probability of observing
their co-occurrence is the product of the marginal probabilities for
each species, \(P(AB) = P(A)P(B)\). In the other case where there is
some positive strength of association between observing both \(A\) and
\(B\) because this interaction is ``important'' for each species, then
the probability of observation both \(A\) and \(B\), \(P(AB)\), is
greater than \(P(A)P(B)\) as \(P(A)\) and \(P(B)\) are not independent
and instead are positively correlated, i.e.~\(P(AB)> P(A)P(B)\). In this
case, the probability of observing a single false-negative in our naive
model from fig.~\ref{fig:geometric}(A) is \(p_{fn}= 1-P(AB)\), which due
to the above inequality implies \(p_{fn}>1-P(A)P(B)\). This indicates an
increasingly greater probability of a false negative as the strength of
association gets stronger, \(P(AB) \to P(AB) \gg P(A)P(B)\). However,
this still does not consider variation in species abundance in space and
time (Poisot \emph{et al.} 2015). If positive or negative associations
between species structure variation in the distribution of \(P(AB)\)
across space/time, then the spatial/temporal biases induced by data
collection would further impact the realized false negative rate, as the
probability of false negative would not be constant for each pair of
species across sites.

@@

A brief caveat here is that we do not consider false-positives---in
large part false-positives can be explained by misidentification of
species, although this could be a relevant consideration in some cases.
The same logic that we apply to false-negatives could easily be applied
to false-positives, e.g.~that we can be much more confident that an
interaction is a true positive if we have observed it 50 times rather
than only once, and we could similarly model this using the geometric
distribution as in fig.~\ref{fig:geometric}(A). However, because
ecological networks are so sparse, there are far more negatives than
positives in the dataset, and therefore likely to be far more
false-negatives than false-positives in absolute terms.

@@

>>>>

Species interactions drive many processes in evolution and ecology. A
better understanding of species interactions is an imperative to
understand the evolution of life on Earth, to mitigate the impacts of
anthropogenic change on biodiversity (Makiola \emph{et al.} 2020), and
for predicting zoonotic spillover of disease to prevent future pandemics
(Becker \emph{et al.} 2021). At the moment we lack sufficient data to
meet these challenges (Poisot \emph{et al.} 2021), largely because
species interactions are hard to sample (Jordano 2016). Over the past
few decades biodiversity data has become increasingly available through
remotely collected data and adoption of open data practices (Kenall
\emph{et al.} 2014; Stephenson 2020). Still, interaction data remains
relatively scarce because sampling typically requires human observation.
This induces a constraint on the amount, spatial scale, and temporal
frequency of resulting data that it is feasible to collect by humans.
Many crowdsourced methods for biodiversity data aggregation (e.g.~GBIF,
eBird) still relies on automated identification of species, which does
not easily generalize to interaction sampling. There is interest in
using remote methods for interaction sampling, which primarily detect
co-occurrence and derive properties like species avoidance from this
data (Niedballa \emph{et al.} 2019). However, this itself is not
necessarily indicative of an interaction (Blanchet \emph{et al.} 2020).
This is an example of semantic confusion around the word
``interaction''---for example one might consider competition a type of
species interaction, even though it is marked by a lack of co-occurrence
between species, unlike other types of interactions, like trophism or
pollination, which require both species to be together at the same place
and time. Here we consider interaction in the latter sense, where two
species have fitness consequences on one-another if (and only if) they
are in the sample place at the same time. In addition, here we only
consider direct (not higher-order) interactions.

@@

The importance of \emph{sampling effort} and its impact on resulting
ecological data has produced a rich body of literature.The recorded
number of species in a dataset or sample depends on the total number of
observations (Walther \emph{et al.} 1995; Willott 2001), as do estimates
of population abundance (Griffiths 1998). This relationship between
sampling effort and spatial coverage and species detectability . This
has motivated more quantitatively robust approaches to account for error
in sampling data in many contexts: to determine if a given species is
extinct (Boakes \emph{et al.} 2015), to determine sampling design (Moore
\& McCarthy 2016), and to measure species richness across large scales
(Carlson \emph{et al.} 2020). In the context of interactions, an initial
concern was the compounding effects of limited sampling effort combined
with the amalgamation of data (across both study sites, time of year,
and taxonomic scales) could lead any empirical set of observations to
inadequately reflect the reality of how species interact (Paine 1988) or
the structure of the network as a whole (Martinez \emph{et al.} 1999;
McLeod \emph{et al.} 2021). Martinez \emph{et al.} (1999) showed that in
a plant-endophyte trophic network, network connectance is robust to
sampling effort, but this was done in the context of a system for which
observation of 62,000 total interactions derived from 164,000
plant-stems was feasible. In some systems (e.g.~megafauna food-webs)
this many observations is either impractical or infeasible due to the
absolute abundance of the species in question.

@@

Here, we illustrate how our confidence that a pair of species never
interacts highly depends on sampling effort. We suggest that surveys of
species interactions can benefit from simulation modeling of the
sampling process. We demonstrate that the realized false-negative rate
of interactions is directly related to the relative abundance of the
species pool, and demonstrate how simulation can be used to produce a
null estimate of the false-negative rate given total sampling effort
(the total count of all individuals of all species seen). We also
introduce a method for resampling of model predictions of interaction
probability to account for observation error. We show that positive
associations in co-occurrence data can increase the realized number of
false-negatives, and demonstrate these positive associations are rampant
in network datasets. We conclude by recommending that the simulation of
sampling effort and species occurrence can and should be used to help
design surveys of species interaction diversity (Moore \& McCarthy
2016), and by advocating use of null models like those presented here as
a tool for both guiding design of surveys of species interactions and
for modeling detection error in predictive models.

@@

\hypertarget{resampling-interaction-probabilities-to-account-for-uncertainty-in-detection-error}{%
\section{Resampling interaction probabilities to account for uncertainty
in detection
error}\label{resampling-interaction-probabilities-to-account-for-uncertainty-in-detection-error}}

Here we show how to incorporate imperfect detection (both
false-negatives and false-positives) into model predictions of
interaction probability. Models for interaction prediction typically
yield a probability of interaction between each pair of species. When
these are considered with uncertainty, it is usually model-uncertainty,
e.g.~the variance in the interaction probability prediction across
several cross-validation folds, where the data is split into training
and test sets several times.Here we introduce a method for resampling
interaction probabilities (outlined in figure fig.~\ref{fig:resampling})
that simulates the observation process with prior estimates of both
false-negative and false-positive probabilities to produce an output
distribution of interaction probability that accounts for observation
error. We implement this in the software package
InteractionUncertaintySampler.jl.

\begin{figure}
\centering
\includegraphics{./figures/uncertainty_sampler.png}
\caption{FIG CAPTION}
\end{figure}

We do this by using the process for simulating a distribution of null
false-negative rates for a given dataset as described above
(fig.~\ref{fig:resampling_concept} (A)). We then consider the output
prediction from an arbitrary prediction model, which is the probability
\(p_{ij}\) that two species \(i\) and \(j\) interact. To get an estimate
of \(p_{ij}\) that accounts for observation error, we resample the
probability of each interaction \(p_{ij}\) by simulating a series of
particles. Each particle is the product of the resampling algorithm
(``Resampling'' within fig.~\ref{fig:resampling:concept} (B)), where the
value of a particle is first drawn from a Bernoulli with weight
\(p_{ij}\), and if that value is true, the value remains true with
probability \(1-p_{fn}\), and if the first draw is false, the value
remains false with probability \(1-p_{fp}\). Over many samples of
particles, the resulting frequency of `true' outcomes is a single
resample of the interaction probability \(p_{ij}^*\). Across many
resamples, this forms a distribution of probabilities which are adjusted
by the true and false negative rates.

As an example case study, we use a boosted-regresssion-tree to predict
interactions in a host-parasite network (Hadfield \emph{et al.} 2014)
(with features derived in the same manner as Strydom \emph{et al.}
(2021) derives features on this data) to produce a set of interaction
predictions. We then applied this method to a set of a few resampled
interaction probabilities between mammals and parasite species shown in
figure fig.~\ref{fig:resampling_concept} (C). Here we implement a simple
resampling algorithm, but this could be extended, for example by
adjusting the expected FNR for each pair of species by the relative
abundance of each species. We have implemented a general framework for
this resampling methodology in the Julia package
InteractionUncertaintySampler.jl, which enables flexible choices for
priors on false-negative and false-positive rates.

This method allows us to more effectively represent our uncertainty
about observation error in predictions of species interactions. This
also has implications for what we mean by `uncertainty' in interaction
predictions. A model's prediction can be `uncertain' in two different
ways: the model's predictions may have high variance, or the model's
predictions may be centered around a probability of interaction of
\(0.5\), where we are the most unsure about whether this interaction
exists. A more statistically robust framework for accounting for
different forms of uncertainty in probabilistic interaction predictions
seems a necessary goal that the field must undertake.

@@

The model above doesn't consider the possibility that there are positive
or negative associations which shift the probability of species
cooccurrence together due to their interaction (Cazelles \emph{et al.}
2016). However, here we demonstrate that the probability of observing a
false negative can be higher if there is some positive association in
the occurrence of species \(A\) and \(B\). If we denote the probability
that we observe the co-occurrence of two species \(A\) and \(B\) as
\(P(AB)\) and if there is no association between the marginal
probabilities of observing \(A\) and observing \(B\), denoted \(P(A)\)
and \(P(B)\) respectively, then the probability of observing their
co-occurrence is the product of the marginal probabilities for each
species, \(P(AB) = P(A)P(B)\). In the other case where there is some
positive strength of association between observing both \(A\) and \(B\)
because this interaction is ``important'' for each species, then the
probability of observation both \(A\) and \(B\), \(P(AB)\), is greater
than \(P(A)P(B)\) as \(P(A)\) and \(P(B)\) are not independent and
instead are positively correlated, i.e.~\(P(AB)> P(A)P(B)\). In this
case, the probability of observing a single false-negative in our naive
model from fig.~\ref{fig:geometric}(A) is \(p_{fn}= 1-P(AB)\), which due
to the above inequality implies \(p_{fn}>1-P(A)P(B)\). This indicates an
increasingly greater probability of a false negative as the strength of
association gets stronger, \(P(AB) \to P(AB) \gg P(A)P(B)\). However,
this still does not consider variation in species abundance in space and
time (Poisot \emph{et al.} 2015). If positive or negative associations
between species structure variation in the distribution of \(P(AB)\)
across space/time, then the spatial/temporal biases induced by data
collection would further impact the realized false negative rate, as the
probability of false negative would not be constant for each pair of
species across sites.

@@

@@

We also should address the impact of false-negatives on the inference of
process and causity in community ecology. We demonstrated that in model
food webs, false-negatives do not impact the measure of total trophic
levels until very high FNR (figure fig.~\ref{fig:addedfnr}(C)), although
we cannot generalize this further to other properties. This has
immediate practical concern for how we design what taxa to sample---does
it matter if the sampled network is fully connected? It has been shown
that the stability of subnetworks can be used to infer the stability of
the metaweb paper beyond a threshold of samples
(\textbf{Song2022RapMon?}). But does this extend to other network
properties? And how can we be sure we are at the threshold at which we
can be confident our sample characterizes the whole system?

To try and sample to avoid all false-negatives is a fool's errand.
Species ranges overlap to form mosaics, which themselves are often
changing in time. Communities and networks don't end in space, and the
interactions that connect species on the `periphery' of a given network
to species outside the spatial extent of a given sample will inevitably
appear as false-negatives in practical samples. The goal should instead
be to sample enough to have a statistically robust estimate of the
current state and empirical change over time of an ecological community
at a given spatial extent and temporal resolution, and to determine what
the sampling effort required the describe the current state prior to
sampling.
